Evolutionary rescue by aneuploidy in tumors exposed to anti-cancer drugs

### 2024-06-27 EvolutionaryRescue1.tex
remove red in discussion as UBD approved

We find that aneuploidy can have a significant effect on evolutionary rescue as it reduces the threshold tumor size by at least an order-of-magnitude even when aneuploidy only provides tolerance (\Cref{rescue_prob}).
When the number of cells in the tumor is large enough (i.e., $N\gg N_m^*\approx 4\times10^7$), a resistance mutation will occur in sensitive cells before these cells become extinct (\Cref{rescue_prob}). Therefore, large tumors are certain to be rescued with or without aneuploidy.
However, anti-cancer drugs are often used as adjuvant therapy after resection, in which case the number of cells in the tumor may be below the detection threshold of $\sim10^7$  \citep{bozic2013evolutionary}. In these cases, aneuploidy can have a crucial role in the evolutionary rescue of the tumor and therefore in cancer recurrence.
{\color{red} Indeed, secondary tumors are estimated to cause the majority of cancer-related deaths~\citep{chaffer2011perspective}. The importance of aneuploidy in the evolutionary rescue of secondary tumors is reinforced by the fact that metastases have been shown to have a chromosome missegregation rate two to three orders of magnitude higher then that of primary tumors~\citep{kimmel2023intra}.}

>>>>

We find that aneuploidy can have a significant effect on evolutionary rescue as it reduces the threshold tumor size by at least an order-of-magnitude even when aneuploidy only provides tolerance (\Cref{rescue_prob}).
When the number of cells in the tumor is large enough (i.e., $N\gg N_m^*\approx 4\times10^7$), a resistance mutation will occur in sensitive cells before these cells become extinct (\Cref{rescue_prob}). Therefore, large tumors are certain to be rescued with or without aneuploidy.
However, anti-cancer drugs are often used as adjuvant therapy after resection, in which case the number of cells in the tumor may be below the detection threshold of $\sim10^7$  \citep{bozic2013evolutionary}. In these cases, aneuploidy can have a crucial role in the evolutionary rescue of the tumor and therefore in cancer recurrence.
Indeed, secondary tumors are estimated to cause the majority of cancer-related deaths~\citep{chaffer2011perspective}. The importance of aneuploidy in the evolutionary rescue of secondary tumors is reinforced by the fact that metastases have been shown to have a chromosome missegregation rate two to three orders of magnitude higher then that of primary tumors~\citep{kimmel2023intra}.
